$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-01 paper/paper.tex
Show named and de Bruijn indexed processes side by side

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax}
\label{syntax}
Let $x, y,\ldots$ range over \emph{variables} and $P, Q,\ldots$ over processes.
The syntax of the \picalc{} \cite{Sangio01} is given by the following grammar.
\begin{equation*}
  \begin{split}
    P,Q ::=      & \; \PO       &\text{(terminated process)}  \\
                |& \; (\new{}x)  P   & \text{(restriction)} \\
                |& \; \comp{P}{Q}  & \text{(parallel composition)} \\
                |& \; \recvp{x}{y}. \; P  & \text{(input)} \\
                |& \; \sendp{x}{y}. \; P   &\text{(output)}
  \end{split}
\end{equation*}
Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)P$ creates a new channel $x$ bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recvp{x}{y}.P$ and $\sendp{x}{y}.P$ denote respectively the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.

Scope restriction $(\new{}x)P$ and input $\recvp{x}{y}.P$ both introduce bound names in $P$.
In order to mechanise the \picalc{} syntax in Agda, we need to deal with bound names in continuation processes.
Names are cumbersome to mechanise: inserting a new variable into a context means proving that its name differs from all other variable names in context.
Instead, we use de Bruijn indices \cite{}, where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago --- hence binders no longer introduce names.
\begin{example}
\hfill{}\\
The \picalc{} term \hfill{} $(\new{}x ) (\comp {\recvp{x}{y} . \; \sendp{y}{\texttt{"Hello world!"}} . \; \PO} {(\new{} y) (\sendp{x}{y} . \; \recvp{y}{z} . \; \PO)})$ \\ is represented as \hfill{} $\new{} (\comp {\recv{0} . \; \send{0}{\texttt{"Hello world!"}} . \; \PO} {\new{} (\send{1}{0} . \; \recv{0} . \; \PO)})$.
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.

A variable occurring under $n$ abstractions has $n$ possible variables to refer to.
References outside of that range are meaningless.
It is useful to rule out these nonsensical terms syntactically.
In  \autoref{fig:var} we do so by introducing the indexed family of types $\Var_n$: for all naturals $n$, the type $\Var_n$ has $n$ distinct elements.

\begin{figure}[h]
\begin{mathpar}
  \datatype
  {n : \N}
  {\Var_n : \Set}
  \; \textsc{Var}

@@

  \inferrule
  {x : \Var_n}
  {\suc x : \Var_{\suc n}}
\end{mathpar}
\caption{Types of size $n$}
\label{fig:var}
\end{figure}

@@

\begin{figure}[h]
\begin{mathpar}
  \datatype
  {n : \N}
  {\Process_n : \Set}
  \; \textsc{Process}
\end{mathpar}

\begin{equation*}
\begin{split}
  \Process_n ::=& \; \PO_n                    \\
              |& \; \new{} \Process_{\suc n}          \\
              |& \; \comp{\Process_n}{\Process_n}          \\
              |& \; \recv{\Var_n}{}\Process_{\suc n} \\
              |& \; \send{\Var_n}{\Var_n}\Process_n
\end{split}
\end{equation*}
\caption{Well-scoped grammar for \picalc{} using de Bruijn indices}
\label{fig:syntax}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax}
\label{syntax}
Let $x, y,\ldots$ range over \emph{variables} and $P, Q,\ldots$ over processes.
The syntax of the \picalc{} \cite{Sangio01} is given by grammar in \autoref{fig:syntax-names}.

\begin{figure}[h]
  \begin{subfigure}{1\textwidth}
    \begin{mathpar}
      \datatype
      {n : \N}
      {\Var_n : \Set}
      \; \textsc{Var}

@@

      \inferrule
      {x : \Var_n}
      {\suc x : \Var_{\suc n}}

      \datatype
      {n : \N}
      {\Process_n : \Set}
      \; \textsc{Process}
    \end{mathpar}
  \end{subfigure}
  \begin{subfigure}[t]{.2\textwidth}
    \begin{equation*}
      \begin{split}
        P,Q ::=      & \; \PO                \\ 
                    |& \; (\new{}x)  P       \\ 
                    |& \; \comp{P}{Q}        \\ 
                    |& \; \recvp{x}{y}. \; P \\ 
                    |& \; \sendp{x}{y}. \; P 
      \end{split}
    \end{equation*}
    \caption{Ill-scoped grammar with names}
    \label{fig:syntax-names}
  \end{subfigure}
  \hspace{.1\textwidth}
  \begin{subfigure}[t]{.7\textwidth}
    \begin{equation*}
      \begin{split}
        \Process_n ::=& \; \PO_n               &&\text{(process termination)}   \\ 
        |& \; \new{} \Process_{\suc n}         &&\text{(restriction)}          \\ 
        |& \; \comp{\Process_n}{\Process_n}    &&\text{(parallel composition)} \\ 
        |& \; \recv{\Var_n}{}\Process_{\suc n} &&\text{(input)}                \\ 
        |& \; \send{\Var_n}{\Var_n}\Process_n  &&\text{(output)}                  
      \end{split}
    \end{equation*}
    \caption{Well-scoped grammar with type-level de Bruijn indices}
    \label{fig:syntax}
  \end{subfigure}
  \caption{}
\end{figure}

Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)P$ creates a new channel $x$ bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recvp{x}{y}. \; P$ and $\sendp{x}{y}. \; P$ denote respectively the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.

Scope restriction $(\new{}x) \; P$ and input $\recvp{x}{y}. \; P$ both introduce bound names in $P$.
In order to mechanise the \picalc{} syntax in Agda, we need to deal with bound names in continuation processes.
Names are cumbersome to mechanise: inserting a new variable into a context means proving that its name differs from all other variable names in context.
Furthermore, names do not guarantee that the syntax well-scoped.
Instead, we use de Bruijn indices \cite{}, where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago --- hence binders no longer introduce names.
\begin{example}
  \begin{equation*}
    \begin{split}
      & (\new{}x ) && (\comp {\recvp{x}{y} && . \; \sendp{y}{\texttt{"Hello world!"}} . \; \PO} {(\new{} y) && (\sendp{x}{y} . \; \recvp{y}{z} && . \; \PO)})
      & \text{names}
      \\
      & \new{} && (\comp {\recv{0} && . \; \send{0}{\texttt{"Hello world!"}} . \; \PO} {\new{} && (\send{1}{0} . \; \recv{0} && . \; \PO)})
      & \text{de Bruijn}
      \\
    \end{split}
  \end{equation*}
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.

A variable occurring under $n$ abstractions has $n$ possible variables to refer to.
References outside of that range are meaningless.
It is useful to rule out these nonsensical terms syntactically.
In \autoref{fig:syntax}, we do so by introducing the indexed family of types $\Var_n$: for all naturals $n$, the type $\Var_n$ has $n$ distinct elements.

@@
